---
title: "Home Work 4"
output: github_document
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(gapminder)
library(tidyverse)
library(knitr)
library(ggthemes)
library(reshape2)
```

#General data reshaping and relationship to aggregation

## Activity #3

####1. Compute the mean  life expectancy for all possible combinations of continent and year. 


```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(Mean_lifeExp=mean(lifeExp))
```

####2. Reshape that to have one row per continent and one variable per year.

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(Mean_lifeExp=mean(lifeExp)) %>% 
  spread(year, Mean_lifeExp)
```

####3.This actually looks really ugly so I think I'll do it by one row per year and one variable for each continent. 

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(Mean_lifeExp=mean(lifeExp)) %>% 
  spread(continent, Mean_lifeExp)
```

####4.Use knitr::kable() to make these tables look pretty in your rendered homework.

```{r}
kable(gapminder %>% 
  group_by(continent, year) %>% 
  summarize(Mean_lifeExp=mean(lifeExp)) %>% 
  spread(continent, Mean_lifeExp))
```

Is there a plot that is easier to make with the data in this shape versis the usual form? If so (or you think so), try it! Reflect.
It is not immediately inutitive how to plot this all on one graph. It is easy to say do year and the mean life expectancy for one continent, like here I did it for Africa. I had to change the axis names since it defaulted to just say Africa (our column title) which is really uninformative. 

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(Mean_lifeExp=mean(lifeExp)) %>% 
  spread(continent, Mean_lifeExp) %>% 
  ggplot(aes(x = year,y = Africa)) + 
  geom_point() +
  geom_smooth(alpha=0.5) + 
  labs(x="Year", 
          y="Mean Life Expectancy",
          title="Mean Life Expectancy Africa Over Time")

```

I did realize that one cool thing that you can do with this is plot countries life expectancies against each other. But I'm not sure that it is easy for readers to interpret since this is data by year but the graph in now way tells you that

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(Mean_lifeExp=mean(lifeExp)) %>% 
  spread(continent, Mean_lifeExp) %>% 
  ggplot(aes(x = Asia,y = Africa)) + 
  geom_point() +
  geom_smooth(alpha=0.5) + 
  labs(x="Mean life expectancy of Asia", 
          y="Mean Life Expectancy of Africa",
          title="Mean Life Expectancy of Asia and Africa")

```

I then remembered that I could convey the years by color though. At least that is a less misleading graph. 

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(Mean_lifeExp=mean(lifeExp)) %>% 
  spread(continent, Mean_lifeExp) %>% 
  ggplot(aes(x = Asia,y = Africa)) + 
  geom_point(mapping= aes(colour= year), alpha=0.5, size=2) +
  geom_smooth(alpha=0.5, size=.5, se = FALSE, color= "pink") + 
  labs(x="Mean life expectancy of Asia", 
          y="Mean Life Expectancy of Africa",
          title="Mean Life Expectancy of Asia and Africa")

```

Finally, to get it all on a single I resorted to adding a geom_point for each continent. 
It is really inefficient coding. I feel like this really only works because there are only five continents, if there were more then this would be a nightmare.

```{r}
gapminder %>% 
  group_by(continent, year) %>% 
  summarize(Mean_lifeExp=mean(lifeExp)) %>% 
  spread(continent, Mean_lifeExp) %>% 
  ggplot(aes(x = year )) + 
  geom_point(aes(y = Asia, colour ="Asia")) +
  geom_point(aes(y = Europe, colour ="Europe")) +
  geom_point(aes(y = Africa, colour ="Africa")) +
  geom_point(aes(y = Oceania, colour ="Oceania")) +
  geom_point(aes(y = Americas, colour ="Americas")) +
  labs(x="Year", 
          y="Mean Life Expectancy",
          title="Mean Life Expectancy Across Time")

```

#Join, merge, look up

##Activity #1

Create a second data frame, complementary to Gapminder.
- I created a Nato country dataset which included two variables: country and natyear  (year the country joined Nato)

```{r}
nato <-read_csv("natocountries.csv")
kable(nato)
```

-Explore the different types of joins:



1. First I decided to explore the lef_join function
```{r}
gapminder %>% 
  left_join(nato) 
```

The left join still returns all rows and columns from the gapminder set so we have a bunch of NAs for all the countries that never joined NATO. 

Now say I wanted only the NATO countries. Then I would use `inner_join` to only return the data for countries that are in my NATO data set. 

```{r}
gapminder %>% 
  inner_join(nato)
```

Next I wanted to do the opposite and only see the data for the countries that are not in NATO: 
```{r}
gapminder %>% 
  anti_join(nato)
```


Next I wanted to explore semi_join. 

```{r}
gapminder %>% 
  semi_join(nato)
```

Semi_join returns all rows from x where there are matching values in y, keeping just columns from x. This appears to do the exact same thing as inner_join for this dataset this is because there are no multiple matches between these datasets.

Now that I've gone through the major functions, I wanted to see about joinging by year to better see the difference between semi_join and inner_join. So I first created a new variable in nato database with the year named the same as the gapminder variable. 

```{r}
natoRename <- nato
colnames(natoRename)[colnames(natoRename)=="natyear"] <- "year"
```
 
First is the inner_join funciton: 

```{r}
gapminder %>% 
  inner_join(natoRename, by ="year") 
```
This returns all the data for each country for the years that have a data point in the Nato dataset. 


Next is the semi_join funciton: 

```{r}
gapminder %>% 
  semi_join(natoRename, by ="year")
```
This returns the data for each country for the years that have a data point in the Nato dataset but removes multiple matches. 

This last part was more just for fun since the year variable in the gapminder dataset means something totally different than the one in the nato dataset. 

## Activity #3

First to compare `merge()` with the dplyr joins. 

```{r}
kable(merge(gapminder, nato))
```

This seems to function the same as the inner_join so it only keeps the gapminder data for which there is a data point in the nato frame- so only data for the countries in NATO.

Now to explore `match()` function: 

```{r}
match(gapminder, nato)
```
This first one returns a vector of NAs. This would normally mean that nothing matches. To verify, I then ran a match with my two nato datasets which I know are identical except for a variable name.

```{r}
match(nato, natoRename)
```

Here we get an output of 1 and 2 meaning that natoRename matches positions 1 and 2 of nato. 