---
title: "Gapminder Exploration"
author: "JC"
date: '2017-09-25'
output: github_document
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(gapminder)
library(tidyverse)
library(knitr)
library(ggthemes)
```

## Smell test the data

We can see from the `structure` command that GapMinder is a data.frame. 

```{r, echo=FALSE}
str(gapminder)
```



We see that Gapminder is a tibble diff data frame.

```{r}
class(gapminder)
```

GapMinder has 6 columns (variables) and 1704 rows of data. 
```{r, echo=FALSE}
ncol(gapminder)
nrow(gapminder)
```



Another way to get this information about extent and size is to use the `dim` function. This will give the dimensions in a single line of code which may save you time. 
```{r, echo=FALSE}
dim(gapminder)
```

If you just want the number of variables, you could also use the `str` funciton (demonstrated above). The first line of output tells you the number of observations and variables. The `length` funciton (below) will give the number of variables 

```{r, echo=FALSE}
length(gapminder)
```

Also if you just type in "gampminder" you will get a data preview which will give you all the variables and their names and tell you the number of rows and tell you the type of each variable. 


```{r, echo=FALSE}
gapminder
```


 At the top of each row we can view the data type of each variable: 
|   **Variable**  | **Data Type** |
|-----------------|------------|
| Country         | Factor     |
| Continent       | Factor     |
| Year            | Integer    |
| LifeExp         | Double     |
| Pop             | Integer    |
| GdpPercap       | Double     |


## Explore individual variables

Here I will explore the variables Continent, year and life expectancy
```{r, echo=FALSE}

kable( gapminder %>% 
  select(continent, year, lifeExp)  %>%
 summary())
```

From this output we can see that Continent has 6 possible values with Africa having the most data points and Oceania having the least. The Year variable tells us that the data was collected between 1952 and 2007 with the mean year being 1980, the distribution of years is relatively uniform. The mean life expectancy across regions and years was 59.47 with a range between 23.60 and 82.60. 

To better see the distribution of Life Expectancy, I created a histogram
```{r, echo=FALSE}
hist(gapminder$lifeExp)
```

This distribution shows a negative skew. 


To explore the variable country, I asked for a summary. From this summary, we can see that each country has 12 data points with data being collected from 142 countries. 

```{r, echo=FALSE}
gapminder %>% 
    group_by(country) %>% 
    summarize(n_distinct(year))
```

## Explore various plot types

```{r, echo=FALSE}
ggplot(gapminder, aes(x=year, y=pop)) +
    geom_point()
```

This plot between population and year seems to be quite interesting. This plot seems to reveal that there are two distinct sub populations here. We see that the vast majority of data points have a low population regarless of year but a few data points show a drastic increase in population over just a few years. 

To further explore this, I created a plot of the population over years for China and Canada. Here we can more easily see how population has grown over the years differently across these two countries


```{r, echo=FALSE, message=FALSE, warning=FALSE}
filter(gapminder, country == "China" | country == "Canada" ) %>%
 ggplot( aes(x=year, y=pop)) +
    geom_point(alpha=1, aes(colour= country))
```

To explore the GDP per capital for the entire data set:
```{r, echo=FALSE}
ggplot(gapminder, aes(x = gdpPercap)) + geom_density()
```

For the entire data set, the vast majority of gdpPercap is under 150000. 
As a comparison, I again focused on just data from China: 


```{r, echo=FALSE}
filter(gapminder, country == "China")  %>%
ggplot( aes(x = gdpPercap)) + geom_density()
```

We see the same general trend in China with most gdpPercap being very low but this graph is much easier to read without any of the larger values beyond 5000. 

For my final graph, I explored the relationship between population and continents: 
```{r, echo=FALSE}
ggplot(gapminder, aes(x=continent, y=pop)) +
    geom_boxplot(outlier.colour = "red")+
geom_jitter(position = position_jitter(width = 0.05),alpha = .15)
```

This graph clearly shows that the population within Asia is much larger than the other continents. However, this graph is very hard to read. Since I have already looked a lot at China, let's focus on Eruope and Oceania: 

```{r, echo=FALSE}
filter(gapminder, continent == "Oceania" | continent == "Europe" )  %>%
ggplot( aes(x=continent, y=pop)) +
    geom_boxplot(outlier.colour = "red")
```

We can still see that Europe has a skew in their data towards larger populations but we can also see that for many of the data points Europe and Oceania had similiar reportings of population.


Just as one last exploration I want to see the life expectancy by year but only for records with populations under half a million.
```{r,echo=FALSE}
kable(gapminder %>% 
    filter(continent == "Asia",
           pop <500000) %>% 
    select(year, lifeExp) )
```

This is a bit hard to read so here is a graph: 
```{r, echo=FALSE}
gapminder %>% 
    filter(continent == "Asia",
           pop <500000) %>%  
    select(year, lifeExp) %>%
ggplot( aes(x=year, y=lifeExp)) +
    geom_point(colour="dark blue") + geom_smooth(se = FALSE,method='lm', color="light blue")
```

For the lower populations, we see a clear increase in life expectancy over the years toping out at just over 70 for 1990s. 

I was curious if higher population places in Asia would see a higher trend, perhaps by being in a larger country that may have more resources. Since I'd like to compare between the same time period, I limited my graph to populations over 500 million and to years prior to 1990. 


```{r, echo=FALSE}
gapminder %>% 
    filter(continent == "Asia",
           pop >500000000, year < 1990) %>%  
ggplot( aes(x=year, y=lifeExp)) +
    geom_point(colour="dark blue") + geom_smooth(se = FALSE,method='lm', colour="light blue")
```

This seems to show the same trend but the life expectancy is lower at every data point.

To better directly compare these groups which are under half a million and those over 500 million:
A word of advice here is to check with View after each mutate line (like below) so that you can check what it is doing as you go.  


```{r}
gapminder  %>% 
  mutate(popsize=NA) %>% 
  mutate(popsize = ifelse(pop > 500000000, "large", popsize)) %>% 
  mutate(popsize = ifelse(pop < 500000, "small", popsize)) %>% 
  filter(!is.na(popsize)) %>% 
   filter(continent == "Asia", year < 1990) %>%  
ggplot( aes(x=year, y=lifeExp)) +
  facet_wrap(~popsize) +
    geom_point(colour="dark blue") + geom_smooth(se = FALSE,method='lm', colour="light blue")
```



Perhaps though a better way to compare is to graph year and life expectancy and vary population plot point sizes:

```{r, echo=FALSE}
gapminder %>% 
    filter(continent == "Asia",
            year < 1990) %>%  
ggplot( aes(x=year, y=lifeExp)) +
    geom_point(colour="pink", aes(size=pop))
```

This makes it a lot easier to see that the lower population countries have higher life expectancy. 



However, it this is a huge simplification as it really depends on how you look at the data. For example if we compare 3 continents prior to 1990 and with populations under half a million:
```{r, echo=FALSE}
gapminder %>% 
    filter(pop <500000,year < 1990) %>%  
ggplot( aes(x=year, y=lifeExp)) +
    geom_point(aes(color=pop)) + facet_wrap(~ continent) 
```


These graphs show differences in the relationship between population and life expectancy between continents. 

Here's another way to show the oversimplificaiton of the above graphs. By focusing on just four countries, you can see that here Japan has the highest population but consistently has higher life expectancy. I also found a theme to make this a bit more organized and professional looking. There is a whole package: ggthemes

```{r, echo=FALSE}
gapminder %>% 
    filter(continent == "Asia", country %in% c("Taiwan", "Vietnam", "Japan", "Lebanon"),
            year < 1990) %>%  
ggplot( aes(x=year, y=lifeExp)) +
    geom_point(aes(size=pop, colour=country)) + ggtitle("Final Graph") +theme_minimal() + labs(x = "Year", y= "Life Expectancy")
```
 

## I want to do more
The example code given does not give all the data points for Rwanda and Afghanistan. The trouble here is that we need to pull all the data for these two countries then vectorize the two countries we are interested in. With the incorrect lne R is pulling Rwanda then Afghanistan and skipping every other year, esentially, it pulls only one country for each year rather than both countries.
```{r}
filter(gapminder, country == c("Rwanda", "Afghanistan"))
```
The proper code is here: 
```{r}
filter(gapminder, country %in% c("Rwanda", "Afghanistan"))
```

Alternatively, you could also write the code like this:
```{r}
filter(gapminder, country=="Rwanda" | country == "Afghanistan")
```

