---
title: "Homework 5"
output: github_document
---


```{r, echo=FALSE, warning=FALSE, message=FALSE}
library(gapminder)
library(singer)
library(tidyverse)
library(knitr)
library(ggthemes)
library(reshape2)
library(dplyr)
library(ggplot2)
library(forcats)
library(gdata)
```

I decided to work with the singer dataset for a change :smile:

## Factor Management

**Factorize.** Transform some of the variable in the `singer_locations` dataframe into factors: pay attention at what levels you introduce and their order. Try and consider the difference between the base R `as.factor` and the `forcats`-provided functions.

First I just want to inspect the data: 
```{r}
data("singer_locations")
kable(head(singer_locations))
```

First I tried to factorize city with the `as_factor` used in class (cm012) but that didn't work. This seems to be due to the existance of NA values. (Here I presented the code as a block quote since knitr wont render a file with code errors.)

> sl_fac <-singer_locations %>% 
  mutate(city_factor = as_factor(city))

>Output: Error in mutate_impl(.data, dots) : Evaluation error: `idx` must contain one integer for each level of `f`.

Remove NA's and then run: 

```{r}
singer_locations %>% 
  filter(!is.na(city)) %>% 
  mutate(city_factor = as_factor(city)) 
kable(head(singer_locations %>% 
  filter(!is.na(city)) %>% 
  mutate(city_factor = as_factor(city)) ))
```

Alternatively, base R also comes with `as.factor`: 
```{r}
sl_facB <- singer_locations %>% 
  mutate(city_factor = as.factor(city)) 
kable(head(sl_facB))
```

A third way is to just use `forcats`. This is often advised since we probably don't want to be reordering our factor which can happen with the base r versions. 
```{r}
sl_facF <- singer_locations %>% 
  mutate(city_factor = factor(city))
kable(head(sl_facF))
```

All of these ways to factor appear to give the same results for our dataset so it probaly doesn't matter which one we go with here. 

**Drop 0.** Filter the `singer_locations` data to remove observations associated with the uncorrectly inputed `year` 0. Additionally, remove unused factor levels. Provide concrete information on the data before and after removing these rows and levels; address the number of rows and the levels of the affected factors.

Prior to removal there were 10100 observations 

```{r}
str(sl_facF)
```
 
 Now with removal of year 0: we are left with 10000 observations

```{r}
sl_facFYear <- (filter(sl_facF, year != "0")) 
    str(sl_facFYear)
```
**Second part of the quesiton:** Additionally, remove unused factor levels. Provide concrete information on the data before and after removing these rows and levels; address the number of rows and the levels of the affected factors.

I for my factor city, there are a numbe of NA values. 
```{r}
sl_facF_cityna <- (sl_facFYear %>% 
    filter(!is.na(city)))
  str(sl_facF_cityna)
```
After removing those we can see that were are 4090 observations with 1316 factor levels for city_factor. 

In examining the data though there are a few odd factors like "?, Illinois, 020, 27, 310 Lousiana, 732 New Jersey" I decided to remove these as well. 

```{r}
sl_new <- (sl_facFYear %>% 
filter( city_factor != "?, Illinois")  %>% 
  filter( city_factor != "020") %>% 
  filter( city_factor != "310, Louisiana") %>% 
  filter( city_factor != "732, NEW JERSEY, USA") %>% 
  filter( city_factor != "27")) 
    str(sl_new) 
```

The structure now indicates that there are only 4083 observations. However, it does still list "?, Illinois". Just to check I tried to call up one of the removed items. 
```{r}
sl_new %>% 
filter(city_factor == "?, Illinois") 
```

It did indead remove the value but I guess it still remembers it as a factor that did exist at one point. 

This was the first bit of code I found:
```{r}
sl_new2 <-(droplevels(sl_new)$city_factor)
str(sl_new2)
```
Okay, this code just created a new table but got rid of all my other data... 

I found on [R-bloggers](https://www.r-bloggers.com/r-drop-factor-levels-in-a-dataset/) advice to use `gdata` package to remove factor levels while retaining your other data. This seems to have worked. 
```{r}
sl_new3 <- drop.levels(sl_new) 
str(sl_new3)
```


**Reorder the levels of `year`, `artist_name` or `title`.** Use the forcats package to change the order of the factor levels, based on a principled summary of one of the quantitative variables. Consider experimenting with a summary statistic beyond the most basic choice of the median.

I decided to reorder `artist_name` by maximum of song duration but since this is so much data and the next step is a graph - I filtered to only 1958. First I had to factorize `artist_name`, then create a summarized coloumn of duration maximum for each artist name. But then I ran into an issue that you can't factor reorder a grouping variable. To solve this I took my newly created `sl_summary` which had the summary statistic that I wanted and left joined it with `sl_new3`

```{r}
sl_small <- (sl_new3 %>% 
  filter(year == 1958))
sl_summary <- (sl_small %>% 
                 filter(year == 1958) %>% 
  mutate(artist_name_factor = factor(artist_name)) %>% 
  group_by(artist_name_factor, song_id) %>% 
  summarize(MaxDur=max(duration))) 
sl_join <- left_join(sl_small, sl_summary)
sl_order <- (sl_join %>% 
               mutate(artist_name_factor = fct_reorder(artist_name_factor, MaxDur )))
levels(sl_order$artist_name_factor)
```


This seems to have worked perfectly. 




